\section{Conclusion}
\label{sec:conclusion}

% In this paper, we looked at problem P with this context and these
% constraints. We proposed solution S. It has such good points and such not so
% good ones. 

Wollok is an educative, object-oriented programming language which is accompanied by an advanced programming environment.
Both tools are highly customized to give support to an introductory OOP course.
Our approach consists of the combination of these three cornerstones: incremental learning path, customized language and specialized programming environment.

First, we defined an \emph{incremental\np{Revisar la idea de Javi} learning path} choosing exactly which are the concepts we want to teach and the order we want to teach them.
The learning path starts with a \emph{simplified programming model} (SPM), \ie one which uses less concepts than a full-fledged OOP language.
The SPM allows the student to build simple programs without requiring more advanced concepts.
The path should attach the SPM with a good set of programming exercises, specifically oriented to be easy to build in the selected SPM.
Once the student has mastered the concepts on the SPM, we can go a step further and introduce the next set of concepts.
\np{Acá se podría hablar de constructivismo.}

Next, defining our own programming language, allows us to give full support to the selected learning path, 
avoiding the need of explaining complex concepts too soon in the course or forcing the student to write \emph{boilerplate code} which he cannot yet understand.
Finally, a good programming environment, helps detecting errors, provides guidance and most significantly allows the student to \emph{explore}.
We have found that often students are afraid to search for solutions not seen in the class or test their own ideas, 
which leads them to restricting themselves into a smaller set of concepts and tools they feel more secure about.
A controlled environment empowers students to look around and explore new possibilities.

\section{Future Work}
% Now we could do this or that.
\label{sec:furtherWork}
One continuous focus of attention for the Wollok IDE development team are the detection programming errors, bad practices, quick fixes, content assistance and refactorings.
A cornerstone to achieve these goals is the type inferer, which is one of our current main objectives.
Still, providing a type inferer for a language such as Wollok has many subtleties, which deserves an independent study \cite{passerini_nicolas_extensible_2014}.
The other half of our future work in this area is a powerfull \emph{effect system} \cite{nielson_type_1999}.

Also, we intend to implement several improvements to the REPL, which we expect to have great impact in the programming experience.
In the first place we want to propagate several of the features of the basic editor to the REPL, such as content assistance.
Then, we would like to enable code modifications while a program is running in the REPL.
%Still, to be effective for novices, we require to develop a strategy to impact changes in code that might have been executed and will not be executed again, such as a variable initializer: in a naïve implementation, changing it would not affect an already instantiated object.
%This kind of \emph{hot} changes can confuse even experienced programmers, so we think it is imperative to think of a specific strategy which beginners can take advantage of.
% ... referencia a scala Worksheets. https://github.com/scala-ide/scala-worksheet/wiki/Getting-Started
After a reload, the REPL should re-execute all the previous expressions in the REPL and display the new results (\cf Scala Worksheets\footnote{\url{https://github.com/scala-ide/scala-worksheet}}).
Finally, we would like an automatic conversion from a REPL session to suite of tests.

Another characteristic of programming in the real world is the need to work in teams. 
The success of object-oriented languages is partly due to their advantages in group projects. 
It is necessary teach our students about the techniques needed for teamwork, right from the beginning. 
To do this, it is essential that the environment has some form of support for group work \cite{kolling_problem_1999}.
Therefore, we plan to create simplified tools to integrate wollok te \emph{version control systems}.

Also there are some initiatives to build web-based or lighter versions of the IDE and the interpreter.
This will allow Wollok to be used in context where the availability of powerful computers is restricted.
There exists a (limited) Wollok web editor integrated to the Mumuki\footnote{\url{http://mumuki.io/}} platform, which yet has not been fully tested with students.
