\section{Introduction}
\label{sec:intro}

% Context
% (a) hay que arrancar introduciendo nuestra visión, contando por qué enseñamos objetos de determinada manera, por qué hicimos Ozono.


% La importancia de arrancar con objetos.
%\emph{Object-oriented programming} (OOP) has become the \textit{de facto} standard programming paradigm in industrial software development.
%Therefore, in the last years software engineering curricula have put more emphasis in object-oriented courses.
% Furthermore, opinion in the community is divided as if it should be the first programming model to be presented. 
% Problem
%Still, students often have difficulties in learning how to program in an object-oriented style which show up both in academic and in industrial environments.

%- aprender a programar se ha revelado como una tarea difícil.
Teaching how to program has revealed itself a difficult task~\cite{dijkstra_89a, jenkins2002difficulty}.
%- detectamos en particular tres aspectos presentes en muchos cursos iniciales de programación que entorpecen (hinder) el aprendizaje: uso de un lenguaje complejo, muchos conceptos para primer programa, entornos no pensados para las necesidades específicas.
As a result of our experience teaching OOP, we have individualized three specific aspects present in many initial programming courses that hinder  the learning process: 
complex programming languages,
too much concepts needed for a first working program, and
programming environments that are not conceived for the specific needs of an initial student~\cite{singh2012}.

%First, OO courses tend to focus too much on syntax and the particular characteristics of a language, instead of focusing on OOP distinctive characteristics.
%Second, many OO languages used in introductory courses do require to grasp a lot of quite abstract concepts before being able to build a first program.
%Finally, poor programming environments are used, although we are at a time where an unexperienced programmer could be making great use of the guidance a good programming environment could provide. These problems are not exclusive of OO courses, they are present in all the general programming courses.

%- múltiples iniciativas de entornos y lenguajes pedagógicos, para distintos públicos y con distintos objetivos.

\medskip 

% Known tracks for solutions
% here you want to show that you are not an idiot not knowing what have been around

%- en particular en objetos hay iniciativas, tanto de presentar un modelo inicial simplificado, como de definir lenguajes y entornos de propósito pedagógico.

% Primero hablar de lenguajes específicos para enseñar
\input{relatedWorks}

% Object first
The great differences between these programming languages and environments show that they have to be analysed in the light of the pedagogical approaches behind them.
The tools are of little use without their respective pedagogical view.
For example, some educational languages and environments are designed to be used in \textit{object-first} approaches, 
\ie for students without any previous programming knowledge \cite{arnow_introduction_1998, bruce_library_2001}.

\medskip
% Nuestro trabajo
%- en nuestra experiencia dando cursos iniciales de OOP, adoptamos un modelo simplificado inicial que prescinde de la noción de clase. Después de algunas semanas, pasamos al modelo clásico de OOP, basado en clases y herencia. Diseñamos e implementamos una serie de entornos iniciales para soportar este modelo que utilizamos en la primer parte del curso. Estos entornos eran adaptaciones del lenguaje y la herramienta principal de edición de Smalltalk, que es el lenguaje que usábamos para el resto del curso. Nos fue rebien (este párrafo está en la intro actual).
Previous works from our team \cite{lombardi_instances_2007,lombardi_carlos_alumnos_2008,griggio_programming_2011,spigariol_lucas_ensenando_2013} have described an approach consisting of
(a) a novel path to introduce OO concepts, focusing on objects, messages, references and polymorphism first, while delaying the introduction of classes and inheritance and 
(b) a reduced and graphical programming environment which supports the order in which we introduce the concepts, by allowing to build OO programs without the need of classes.
This way of organizing a course provides a more gentle learning curve to students and allows them to write completely working programs from the first lectures.

%- A pesar del éxito, los entornos que implementamos no nos resultaron satisfactorios en algunos aspectos entre los que mencionamos: 
While this approach proved to be successful in providing the students with a more profound knowledge of OOP at the same time as raising pass rates, 
we feel that there is still room for improvement, in three areas:
%  a. la gran distancia entre la experiencia en la materia y la realidad de la industria, tanto en el lenguaje como en las herramientas de desarrollo utilizadas
(a) the difference between the experience in the classroom and the reality in (most) professional environments, both in the language as in the development tools.
%  b. el salto entre el modelo simplificado y el modelo clásico que se daba en el curso, principalmente por la diferencia en las herramientas de desarrollo.
(b) the gap between the simplified programming model and the classical model, mostly because of the differences between the development tools
%  c. atarnos a un lenguaje industrial, que nos limita para definir con precisión las herramientas pedagógicas.
(c) adherence to an industrial language limited pedagogical decisions

% What our solution is \ct{Set} and \ct{OrderedCollection} (so that the reader knows where the paper is going)
% Contribution of the paper
% (b) lo que aprendimos en estos 8 años haciendo eso, que nos lleva a querer darle una vuelta más.
%As a result, we decided to conceive both a programming language and an accompanying development environment, that follows closely following the pedagogical approach we advocate for an initial OOP course. 
%The main goal of this paper is to describe Wollok\footnote{
	%\url{http://www.wollok.org/}. 
	%Source code and documentation can be found in Github 
	%(\url{https://github.com/uqbar-project/wollok}).
	%Wollok is open-source and distributed under LGPLv3 License 
	%(\url{http://www.gnu.org/copyleft/lgpl.html}).}, 
%the tool we created that reunites language and environment. We point out the pedagogical considerations that suggested us to build Wollok, and how they influence various of its design decisions. We also report briefly our two-year experience using this tool in initial OOP courses.

\medskip
The main goal of this paper is to describe Wollok\footnote{
	\url{http://www.wollok.org/}. 
	Source code and documentation can be found in Github 
	(\url{https://github.com/uqbar-project/wollok}).
	Wollok is open-source and distributed under LGPLv3 License 
	(\url{http://www.gnu.org/copyleft/lgpl.html}).}, 
the teaching tool we conceived to cope with the beforehand considerations.
Wollok reunites an ad-hoc programming language with an accompanying development environment. 

The design of both components of Wollok follows closely the pedagogical approach we advocate for an initial OOP course. 
Namely, Wollok combines some characteristics that are typically present in academical environments with others that are more easily seen in industrial ones, aiming to enrich them to fulfill the specific needs of novice programmers and the proper development of a first OOP course.
Its most noticeable characteristics are:
(1) the possibility of using self-defined objects to support an introduction to the topic that postpones the introduction of the concept of class,
(2) the possibility of combining self-defined with class-based code objects in the same program,
(3) the decission of offering an IDE with edition and code management capabilities that are fine-tuned for unexperienced programmers
and (4) that both the language as the IDE share similarities with their industrial counterparts, in order to soften the later transition to the professional tools.

\medskip 
% Paper structure
%The rest of the paper is structured as follows. 
\emph{Plan of the paper.}
In \secref{problem} we present the problems of learning Object Oriented programming, and the consequences of this difficulties to the students. 
\secref{wollokLanguage} describes the Wollok language and the design goals and ideas we take in consideration for its design. 
\secref{environment} is devoted to the Wollok integrated development environment, outlying its features and their pertinence for the teaching of programming skills. 
In \secref{experience}, we report briefly the actual experience accumulated to this date using Wollok in several universities.
In \secref{discussion} we outline some of the discussions that lead to the main design decisions behind Wollok.
%, while \secref{related} compares our solution with another similar approaches. 
Finally, in \secref{conclusion} and \ref{sec:furtherWork} we summarize our contributions, along with some possible lines of further work. 
%As an appendix we have Sections \ref{sec:implementation}, \ref{sec:wollokSoftware} and \ref{sec:ChecksAndValidations} which give more details about the implementation.


