\documentclass{article}
\usepackage[T1]{fontenc} %%%key to get copy and paste for the code!
\usepackage[utf8]{inputenc} %%% to support copy and paste with accents for frnehc stuff
\usepackage{times}
\usepackage[scaled=0.85]{helvet}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{xspace}
\usepackage{alltt}
\usepackage{latexsym}
\usepackage{url}            
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage{cite}
\usepackage[pdftex,colorlinks=true,pdfstartview=FitV,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{xspace}

\input{macros}

\begin{document}
\title{Wollok -- y una frase que describa la idea}
\author{Nicolás Passerini \and Javier Fernandes}
\date{\today}
\maketitle

\begin{abstract}
In this context...
We consider this problem P...
P is a problem because...
We propose this solution...
Our solution solves P in such and such way.
\end{abstract}


\section{Introduction}
\label{sec:intro}

% Context
% (a) hay que arrancar introduciendo nuestra visión, contando por qué enseñamos objetos de determinada manera, por qué hicimos Ozono.

Enseñanza tradicional:
- Muchas veces se parte del lenguaje
- Bajos niveles de aprobación
- Pasa mucho tiempo hasta que un chico puede hacer un programa "real", demasiados conceptos.
- ... después pienso más

Nosotros propusimos:
- Elegir un recorrido que permite ir incorporando los conceptos uno a uno
- Tener una herramienta que da soporte a eso
- Focalizar en objeto-mensaje-polimorfismo-referencias, los demás conceptos aparecen después.


Also these hindrances reduce the opportunity of students to apply
the concepts of the paradigm effectively in their further
professional practice, resulting in several IT-projects not taking
advantage of the possibilities offered by the potential of good
object-oriented practices, even in cases where the tools used may
allow the application of object-oriented programming gracefully. \cite{lombardi_instances_2007}

Secondly, better environments have become necessary. Earlier introductory courses
focused on the development of algorithms in procedural or functional languages. To
do this, an editor and a compiler was all that was needed for the practical part of the
work. Modern courses now use object-oriented languages and subject material
taught includes testing, debugging and code reuse. This creates the need to deal with
multiple source files and multiple program development tools from the very start. To
give a beginning student a chance to cope with this increased complexity, better
environment support is needed. \cite{kolling_problem_1999}

% Problem

\medskip 

% Known tracks for solutions
We propose to provide the student a reduced and graphical
programming environment in which the object and the message are
the central concepts instead of defining classes and then instantiate
them. \cite{griggio_programming_2011}

% here you want to show that you are not an idiot not knowing what have been around


\medskip 

% What our solution is \ct{Set} and \ct{OrderedCollection} (so that the reader knows where the paper is going)

(b) lo que aprendimos en estos 8 años haciendo eso, que nos lleva a querer darle una vuelta más.

Si bien todo esto que se nos ocurrió en su momento fue maravilloso y genial y los pibes aprenden más, programan mejor y la ponen más seguido, fuimos aprendiendo más cosas y estamos trabajando ne unos cambios:

Los fáciles:
- integrar clases y objetos en un mismo programa
- integrar interfaces de usuario automáticas
- utilizar herramientas avanzadas para guiar a los alumnos en el proceso de aprendizaje, la vedette acá sería el sistema de tipos.
- acercar la práctica de lo que hacemos a la práctica industrial (acá hay que ver qué decimos y qué no, por ejemplo me molesta que no sea un archivo... pero desde lo metodológico se puede pensar en tests o incluso en un repositorio de código... sobre la relación con la industria se podría escribir un libro, hay que ver cuánto queremos meternos).

No sé si hablar de los temas de colecciones, son un poco particulares de Ozono.


\medskip 

% Contribution of the paper
(c) las ideas nuevas... y wollok como herramienta para dar soporte a esas ideas.

Lo concreto que hicimos es tener un lenguaje con
- clases y objetos integrados
- un IDE profesional con syntax highlighting, refactors, autocompletion y la vedette (?) inferencia de tipos (en progreso).
- bueno, muchas mejoras a nivel lenguaje, como literales para coleciones, imports
- una forma fácil de construir tests.

\medskip 

% Paper structure


\section{Problem Description}
\label{sec:problem}

% Context, exposed with the \textbf{most precise terms possible} (don't open
% unwanted doors for the reader)

% Probably set the vocabulary before to cut any misinterpretation

% Constraints that influenced the solution (because the solution is not
% universal) \emph{e.g.} our requirements for a solution, possibly not all
% satisfied. They should be sound and believable. Analysis of the criteria.
% Imagine that you are another guy having this problem do the constraint
% matches yours so that you could apply the solution

% Problem

This means students must dive
right into classes and objects, their encapsulation (public
and private data, etc.) and methods (the constructors,
accessors, modifiers, helpers, etc.). All this is in addition to
mastering the usual concepts of types, variables, values, and
references, as well as with the often-frustrating details of
syntax. Now, add event-driven concepts to support
interactivity with GUIs! As argued by [11], learning to
program objects-first requires students grasp "many
different concepts, ideas, and skills…almost concurrently.
Each of these skills presents a different mental challenge." \cite{cooper_teaching_2003}

% Factual solution tracks, to position...

Ozono hereda muchos de los problemas de los entornos Smalltalk

Smalltalk, however, lacks other important facilities: no visualisation tools for class
relations are available. The main problem with this lies in the Smalltalk language
itself: since it is not statically typed, it is not possible to extract usage relations from
its source code. No indication exists before runtime as to the call relationships
between classes. Inheritance relationships as shown in the browser do not present
the relationships of one application but rather the whole Smalltalk environment and
so the browser is not used as an application modelling tool. Smalltalk blurs the
distinction between the environment and the application under development.
Reports about the use of Smalltalk systems for teaching also point to another
problem: its size. While the language itself (in terms of the number of constructs) is
small, the class library and tools are large and often confusing. Several authors
reported difficulties with the students ability to cope with the environment [7, 8],
especially that experimentation and self directed learning was not working well
because students were overwhelmed by the system. They also found that the
functionality of the browser should be limited, since its power and flexibility caused
more problems than it solved. \cite{kolling_problem_1999}.

% Our solution in a nutshell.


\section{Proposed Solution}
\label{sec:contribution}
% Free form, variable number of sections, technical details.
% But in general do not mix solution and discussions/possible variation let that for discussion

\section{Discussion}
\label{sec:discussion}

% Discussion of actual solution \emph{vs.} initial constraints from \ref{sec:problem}. Explain the space of the solution, why we made it this way.

% Evaluation of the solution. How does the solution meet the criteria? Where does it succeed or fails...


\section{Related Works}
\label{sec:related}

% Other solutions in the domain, and a real comparison of our contribution with solutions from other people.
% Base tomada del paper de ESUG 2011
LOOP is presented as a visual environment to teach OOP using
a reduced set of language constructions and a prototype approach
to create objects. It presents the main concepts of object, message
and reference in a specialized tool with a visual representation of
the object environment. Several visual tools to teach programming
already exists, like ObjectKarel[2], Scratch[14] and Etoys[4].
ObjectKarel presents a visual tool based on the abstraction of
robots to teach OOP, using a map where the robots-the objects-
move when messages are sent to them. LOOP does not center on a
specific abstraction like a robot: it allows the student to create any
other abstraction. Scratch and Etoys, are aimed to teach the basics
of programming to children, using visual objects and scripts to play
tween objects. This kind of diagrams could be inferred from the
evaluation of any piece of code, even the execution of tests.
Another subject of research is a “debugger” for the tool [1]. We
think that a live and powerful debugger a ` la Smalltalk is a rich tool
for the understanding of the whole environment behaviour. After a
message is sent, a debugger view can be used like a video player,
with play, forward and backward buttons to navigate the message
stack and see how the state changes after each message send in the
object diagram.
Finally, there are some improvements to be made to the user
interface, such as shortcuts, code completion, improved menus or
internationalization. Currently the tool is only available in spanish,
we want to make it configurable to add more languages as neces-
sary.

% Agregar referencia al paper de Fidel, y otros

\section{Conclusion}
\label{sec:conclusion}

% In this paper, we looked at problem P with this context and these
% constraints. We proposed solution S. It has such good points and such not so
% good ones. 

% Now we could do this or that.

Y a futuro le agregaríamos:
- Integración con la UI
- Integración fácil con SCM

Another characteristic of programming in the real world is the need to work in
teams. The success of object-oriented languages is partly due to their advantages in
group projects. Ideally, we also want to teach our students about the techniques
needed for teamwork. To do this, it is essential that the environment has some form
of support for group work. \cite{kolling_problem_1999}

- Más refactors y mejora del sistema de inferencia.
- Una versión web / versión liviana con el objetivo de poder ejecutarse en las netbooks que tienen los chicos de secundaria.

Probar la nueva herramienta en entornos educativos.

Eso pensando el lenguaje/herramienta, si pienso a nivel docencia se me ocurre que lo que tenemos que hacer es integrarnos con otras entidades, como Sadosky u otras universidades.
Haciendo foco en que el punto no es la herramienta sino que tenemos que repensar cómo enseñamos a programar.

\subsection*{Acknowledgements} 
% This work was supported by Ministry of Higher Education and Research, Nord-Pas de Calais Regional Council, FEDER through the 'Contrat de
% Projets Etat Region (CPER) 2007-2013',  the Cutter ANR project, ANR-10-BLAN-0219 and the MEALS Marie Curie Actions program FP7-PEOPLE-2011-
% IRSES MEALS (no. 295261). 

% \bibliographystyle{plain}
% \bibliography{foo.bib}

% \appendix
% 
% \section{Lots of Furry Technical Details}

\bibliographystyle{abbrv}
\bibliography{wollok,teaching,scg}

\appendix
\section{Images}
Imágenes y otros detalles de wollok que no entran en las 6/7 páginas del artículo
\end{document}
